Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-09-07 concurrency/CPP 2024/paper.tex
fix breakage

\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n'.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure? Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations. However, in most sequential settings ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself. The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to data structures without modifying them. Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t}:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {
		node *t; 
		lock_t *lock; 
		int min; 
		int max;
	} node_t;
\end{lstlisting}
These fields store lower and upper bounds on the keys reachable from the current node. %In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it. %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
Our specification of \lstinline{inRange} is:
\begin{mathpar}
	% inRange for give-up
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2
		\end{array}
		\right\}
	}
	\vspace{-6pt}\\ \vspace{-6pt}
	\texttt{inRange(pn, k)} 
	\\
	{\color{blue}
		\left\{\mathit{res.} \ 
		\begin{array}{l}
			\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2 \ \ast  \ 
			\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2))
		\end{array}
		\right\}
	}
\end{mathpar}
It simply computes whether the input key \lstinline{k} is in the range $(n_1, n_2)$ associated with the node \texttt{n}, and returns 0 or 1 accordingly.

>>>>

\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure? Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations. However, in most sequential settings ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself. The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to data structures without modifying them. Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t}:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {
		node *t; 
		lock_t *lock; 
		int min, max;
	} node_t;
\end{lstlisting}
These fields store lower and upper bounds on the keys reachable from the current node. %In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it. %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
Our specification of \lstinline{inRange} is:
\begin{mathpar}
	% inRange for give-up
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2
		\end{array}
		\right\}
	}
	\vspace{-6pt}\\ \vspace{-6pt}
	\texttt{inRange(pn, k)} 
	\\
	{\color{blue}
		\left\{\mathit{res.} \ 
		\begin{array}{l}
			\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2 \ \ast  \ 
			\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2))
		\end{array}
		\right\}
	}
\end{mathpar}
It simply computes whether the input key \lstinline{k} is in the range $(n_1, n_2)$ associated with the node \texttt{n}, and returns 0 or 1 accordingly.
